---
title: "Route Graph"
output:     
    MetroTransitr::metro_html: default
    MetroTransitr::metro_report: default
---

<style type="text/css">
html {
  max-width: 10.5in;
  margin: 1.25in 0.75in;
  font-family: "Calibri", helvetica, sans-serif;
  font-size: 11pt;
}

body { width: 8.5in;
       max-width: 10.5in;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(data.table)
library(igraph)
library(sp)
library(RColorBrewer)
library(leaflet)

OD <- readRDS('data/OD.RDS')
setDT(OD)

tracts <- readRDS('data/tracts.RDS')

OD_graph <- graph_from_data_frame(OD[, .(Var1, Var2, weight)], directed = FALSE)
```

## Descriptive Statistics
### Degree
The degree of a node is the number of edges connected to that node, and is a basic measure of connectedness. Below is a map showing tracts with higher degree in darker red. On average, tracts have degree 112.7 with median 84 degrees.
```{r}
tot_degrees <- degree(OD_graph, loops = TRUE, mode = 'total')
tot_degrees <- data.table(tract = names(tot_degrees), tot_degree = tot_degrees)
tract_deg <- merge(tracts, tot_degrees, by.x = 'GEOID10', by.y = 'tract')

pal <- colorNumeric(palette = "Reds", domain = tract_deg$tot_degree)

leaflet() %>% addTiles(urlTemplate = 'http://tile.stamen.com/toner-lite/{z}/{x}/{y}.png',
                       attribution = 'Map tiles by <a href="http://stamen.com">Stamen Design</a>,
                       under <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a>.
                       Data by <a href="http://openstreetmap.org">OpenStreetMap</a>,
                       under <a href="http://www.openstreetmap.org/copyright">ODbL</a>.') %>%
  setView(lat = 44.963, lng = -93.22, zoom = 11) %>%
  addPolygons(data = tract_deg, fillColor = ~pal(tot_degree), fillOpacity = .7, weight = 1)
```


We can also compare to the weighted degree, defined for a node as the sum of all edge weights for edges connected to that node. The average weighted degree is 717.0 with median 171. One can think of the unweighted degree as information about the number of other tracts a given tract is connected to, and the weighted degree as the number of possible routes to or from a tract.
```{r}
wght_degrees <- strength(OD_graph, loops = TRUE, mode = 'total')
wght_degrees <- data.table(tract = names(wght_degrees), wght_degree = wght_degrees)
tract_deg <- merge(tracts, wght_degrees, by.x = 'GEOID10', by.y = 'tract')

pal <- colorNumeric(palette = "Reds", domain = tract_deg$wght_degree)

leaflet() %>% addTiles(urlTemplate = 'http://tile.stamen.com/toner-lite/{z}/{x}/{y}.png',
                       attribution = 'Map tiles by <a href="http://stamen.com">Stamen Design</a>,
                       under <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a>.
                       Data by <a href="http://openstreetmap.org">OpenStreetMap</a>,
                       under <a href="http://www.openstreetmap.org/copyright">ODbL</a>.') %>%
  setView(lat = 44.963, lng = -93.22, zoom = 11) %>%
  addPolygons(data = tract_deg, fillColor = ~pal(wght_degree), fillOpacity = .7, weight = 1)
```

### Degree Distribution

Below is a plot of the (empirical) cdf of the unweighted degrees. Nearly all of the tracts have degree < 300, although the maximum degree is 2710.
```{r}
tot_degrees[, N := .N, by = .(tot_degree)]
tot_degrees[, pct := N/583]
unq <- unique(tot_degrees[, .(tot_degree, pct)], by = 'tot_degree')
unq <- unq[order(tot_degree)]
unq[, cum_sum := cumsum(pct)]

plot(unq[, tot_degree], unq[, cum_sum])
```

Here is a plot of the (empirical) cdf of the weighted degrees.
```{r}
setkey(wght_degrees, wght_degree)
#Compute number of tracts per each possible weighted degree
wght_degrees[, N := .N, by = wght_degree]

#Compute percentage for each possible weighted degree
wght_degrees[, pct := N/583]

#Get unique weighted degrees
unique_degs <- unique(wght_degrees[, .(wght_degree, pct)], by = 'wght_degree')

#Find cumulaive sum = empirical cdf
unique_degs[, cum_sum := cumsum(pct)]

plot(unique_degs[, wght_degree], unique_degs[, cum_sum])

```

###Clustering Coefficient
The clustering coefficient of a node measures how close a node and its neighbors are to forming a complete graph. A complete graph is a graph where all possible edges are present. The clustering coefficient is given as the proportion of edges connected to a node compared to the total number of possible edges between the node and its neighbors. Below is a map showing the clustering coefficient of the tracts.

It may initially be surprising that downtown Minneapolis and St. Paul have low clustering coefficients, and many outlying areas have high clustering coefficients. However, a tract may have high degree, but tracts at the other 'ends' of the edges are not themselves connected leading to a low clustering coefficient. This may help explain the phenomenon of the low clustering coefficient of downtown Minneapolis, downtown St. Paul and the airport/Mall of America. This may mean that Minneapolis, St. Paul and the airport are well-connected, but the outlying areas are not connected to one-another.
```{r}
clust_coeff <- transitivity(OD_graph, type = 'local')
clust_coeff <- data.table(tract = names(V(OD_graph)), clust_coeff = clust_coeff)

tract_cc <- merge(tracts, clust_coeff, by.x = 'GEOID10', by.y = 'tract')

pal <- colorNumeric(palette = 'Reds', domain = tract_cc$clust_coeff)

leaflet() %>% addTiles(urlTemplate = 'http://tile.stamen.com/toner-lite/{z}/{x}/{y}.png',
                       attribution = 'Map tiles by <a href="http://stamen.com">Stamen Design</a>,
                       under <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a>.
                       Data by <a href="http://openstreetmap.org">OpenStreetMap</a>,
                       under <a href="http://www.openstreetmap.org/copyright">ODbL</a>.') %>%
  setView(lat = 44.963, lng = -93.22, zoom = 11) %>%
  addPolygons(data = tract_cc, fillColor = ~pal(clust_coeff), fillOpacity = .7, weight = 1)
```

## Betweenness Centrality

The Betweenness Centrality of a node measures how 'central' that node is to the graph. Specifically, the betweenness centrality of a node k is given by
\[
BC(k) = \sum_{(i,j) \in E} \delta_{i,j}(k)/\delta_{i,j},
\]
where $\delta_{i,j}$ is defined as the total number of shortest paths between nodes $i$ and $j$, and $\delta_{i,j}(k)$ is the number of shortest paths between nodes $i$ and $j$ that pass through node $k$. Intuitively, betweenness centrality assumes that a traveler on a graph will take the shortest path between any two points, and a higher betweenness centrality means more travelers pass through that node.

The map below shows the relative betweeness centrality of census tracts when the OD data is considered as an undirected graph. Downtown 
Minneapolis, downtown St. Paul, and the airport/Mall of America all have high betweenness centrality meaning many shortest paths pass through these census tracts.

```{r}
bet_tract <- betweenness(OD_graph, weight = NULL, directed = FALSE)
bet_tract <- data.table(tract = names(bet_tract), BC = bet_tract)
tracts <- merge(tracts, bet_tract, by.x = 'GEOID10', by.y = 'tract', all.y = TRUE)

pal_tract <- colorNumeric(palette = "Reds", domain = tracts$BC)
tract_BC_leaf <- leaflet() %>% addTiles(urlTemplate = 'http://tile.stamen.com/toner-lite/{z}/{x}/{y}.png',
                            attribution = 'Map tiles by <a href="http://stamen.com">Stamen Design</a>,
                            under <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a>.
                            Data by <a href="http://openstreetmap.org">OpenStreetMap</a>,
                            under <a href="http://www.openstreetmap.org/copyright">ODbL</a>.') %>%
  setView(lat = 44.963, lng = -93.22, zoom = 11) %>%
  addPolygons(data = tracts, fillColor = ~pal_tract(BC), fillOpacity = .7, weight = 1)

tract_BC_leaf
```

## Page Rank
Page Rank was an algorithm originally proposed to filter websites for a search engine. A website has a higher Page Rank if a 'random clicker' is more likely to end up on that site. The same method can be applied to any graph. In the OD graph, we can think Page Rank as measuring destinations in the sense that a higher Page Rank means a traveler is more likely to end their trip in that census tract.

In the graph below, it is evident that Minneapolis is the most likely trip destination. While St. Paul and the airport were strong hubs (many trips passing through the location), they were less strong centers (many trips ending at the location).

```{r}
pr_tract <- page_rank(OD_graph, vids = V(OD_graph), weights = edge_attr(OD_graph, name = 'weight', index = E(OD_graph)))
pr_tract <- data.table(tract = names(pr_tract$vector), PR = pr_tract$vector)
tracts <- merge(tracts, pr_tract, by.x = 'GEOID10', by.y = 'tract', all.y = TRUE)

pal_tract_PR <- colorNumeric(palette = "Reds", domain = tracts$PR)
tract_PR_leaf <- leaflet() %>% addTiles(urlTemplate = 'http://tile.stamen.com/toner-lite/{z}/{x}/{y}.png',
                       attribution = 'Map tiles by <a href="http://stamen.com">Stamen Design</a>,
                       under <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a>.
                       Data by <a href="http://openstreetmap.org">OpenStreetMap</a>,
                       under <a href="http://www.openstreetmap.org/copyright">ODbL</a>.') %>%
  setView(lat = 44.963, lng = -93.22, zoom = 11) %>%
  addPolygons(data = tracts, fillColor = ~pal_tract_PR(PR), fillOpacity = .7, weight = 1)

tract_PR_leaf
```

## Random Walk and Walktrap Clustering
A random walk on a graph creates a path, where the next edge in the path is chosen randomly. Starting at node $i$, the probability that the random walker travels to node $j$ is proportional to the weight of edge $(i,j)$. For example if edge $(i,j)$ has weight $w_{ij} =1$ and edge $(i,k)$ has weight $w_{ik} = 2$, the random walk is more likely to choose edge $(i,j)$ as the next part of the path.

Below are graphs for the clusters generated by a random walk of lengths 6 and 10 respectively.

```{r}
tract_wt <- cluster_walktrap(OD_graph, steps = 6, weights = E(OD_graph)$weight)
#tract_wt
tract_wt_memb <- membership(tract_wt)
tract_wt_memb <- data.table(tract = names(tract_wt_memb), clust = tract_wt_memb)
tract_wt <- merge(tracts, tract_wt_memb, by.x = 'GEOID10', by.y = 'tract', all.x = TRUE)

pal <- colorFactor(palette = "Paired", domain = tract_wt$clust)
leaflet() %>% addTiles(urlTemplate = 'http://tile.stamen.com/toner-lite/{z}/{x}/{y}.png',
                       attribution = 'Map tiles by <a href="http://stamen.com">Stamen Design</a>,
                       under <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a>.
                       Data by <a href="http://openstreetmap.org">OpenStreetMap</a>,
                       under <a href="http://www.openstreetmap.org/copyright">ODbL</a>.') %>%
  setView(lat = 44.963, lng = -93.22, zoom = 9) %>%
  addPolygons(data = tract_wt, fillColor = ~pal(clust), fillOpacity = .7, weight = 1)
```

```{r}
tract_wt <- cluster_walktrap(OD_graph, steps = 10, weights = E(OD_graph)$weight)
#tract_wt
tract_wt_memb <- membership(tract_wt)
tract_wt_memb <- data.table(tract = names(tract_wt_memb), clust = tract_wt_memb)
tract_wt <- merge(tracts, tract_wt_memb, by.x = 'GEOID10', by.y = 'tract', all.x = TRUE)

pal <- colorFactor(palette = "Paired", domain = tract_wt$clust)
leaflet() %>% addTiles(urlTemplate = 'http://tile.stamen.com/toner-lite/{z}/{x}/{y}.png',
                       attribution = 'Map tiles by <a href="http://stamen.com">Stamen Design</a>,
                       under <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a>.
                       Data by <a href="http://openstreetmap.org">OpenStreetMap</a>,
                       under <a href="http://www.openstreetmap.org/copyright">ODbL</a>.') %>%
  setView(lat = 44.963, lng = -93.22, zoom = 9) %>%
  addPolygons(data = tract_wt, fillColor = ~pal(clust), fillOpacity = .7, weight = 1)
```